Conversations in time: interactive visualisation to explore structured temporal data

### 2020-09-23 tsibbletalk.Rmd
background section

The \CRANpkg{tsibble} package [@wang2020tsibble] introduces a unified temporal data structure, referred to as a `tsibble`, to represent time series and longitudinal data in a tidy format [@wickham2014tidy]. That said, a `tsibble` extends the `data.frame` and \CRANpkg{tibble} classes with temporally contextual metadata: `index` and `key`. The `index` declares a data column that holds time-related indices. The `key` identifies a collection of related series or panels observed over the `index`-defined period, which can comprise multiple columns. Below displays the monthly Australian retail trade turnover data (`aus_retail`), available in the \CRANpkg{tsibbledata} package [@R-tsibbledata]. The `Month` column holds year-months as `index`. The `State` together with `Industry` are the identifiers for these 152 series, highlighted as `key`. Note that the column `Series ID` could be an alternative option for setting up `key`, but `State` and `Industry` are more readable and informative. The `index` and `key` are "sticky" columns to a `tsibble`, forming critical pieces for fluent temporal data analysis later.

@@

In the spirit of tidy data to the \CRANpkg{tidyverse} [@Wickham2019], the **tidyverts** suite features `tsibble` as the foundational data structure, in order to build a fluid and fluent pipeline for time series analysis. Besides \CRANpkg{tsibble}, the \CRANpkg{feasts} [@R-feasts] and \CRANpkg{fable} [@R-fable] packages fill the role of statistical analysis and forecasting in the **tidyverts** ecosystem. When time series analysis starts taking off, series of interest denoted by the `key` variables often remain unchanged over the course of analysis, from trend inspection to forecasting performance.

Figure \ref{fig:highlight-retail-1} gives an overview of 152 series for the retail data using an overlaid time series plot, while Figure \ref{fig:highlight-retail-2} presents a scatterplot, where each series is represented by a dot in the feature space (trend versus seasonal strength). The plot making of Figure \ref{fig:highlight-retail-2} is aided with the `features()` function from \CRANpkg{feasts}, which summarises original data by each series down to various statistical features. This function along with other **tidyverts** functions is `tsibble`-aware, and outputs a table in a reduced form where each row corresponds to a series, thus graphically displayed as Figure \ref{fig:highlight-retail-2}.

@@

Figure \ref{fig:highlight-retail} highlights not only a series with strongest seasonality, but also a need to querying interesting series on the fly. Without interactivity, one needs to first filter the interesting series out from the features table, and join back to the original `tsibble` in order to examine its trend in relation to others. This procedure can soon grow cumbersome if many series to be discovered. Despite that the two plots are static, they can be considered as linked views via the common `key` variables between two tables. This motivates enabling interactivity of `tsibble` and `tsibble`-derived objects for rapid exploratory data analysis.

>>>>

The \CRANpkg{tsibble} package [@wang2020tsibble] introduced a unified temporal data structure, referred to as a `tsibble`, to represent time series and longitudinal data in a tidy format [@wickham2014tidy]. A `tsibble` extends the `data.frame` and \CRANpkg{tibble} classes with the temporal contextual metadata: `index` and `key`. The `index` declares a data column that holds time-related indices. The `key` identifies a collection of related series or panels observed over the `index`-defined period, which can comprise multiple columns. An example of a `tsibble` can be found in the monthly Australian retail trade turnover data (`aus_retail`), available in the \CRANpkg{tsibbledata} package [@R-tsibbledata], shown below. The `Month` column holds year-months as the `index`. `State` and `Industry`, together, are the identifiers for these 152 series, which form the `key`. Note that the column `Series ID` could be an alternative option for setting up the `key`, but `State` and `Industry` are more readable and informative. The `index` and `key` are "sticky" columns to a `tsibble`, forming critical pieces for fluent downstream temporal data analysis.

@@

In the spirit of tidy data from the \CRANpkg{tidyverse} [@Wickham2019], the **tidyverts** suite features `tsibble` as the foundational data structure, and helps to build a fluid and fluent pipeline for time series analysis. Besides \CRANpkg{tsibble}, the \CRANpkg{feasts} [@R-feasts] and \CRANpkg{fable} [@R-fable] packages fill the role of statistical analysis and forecasting in the **tidyverts** ecosystem. During all the steps of a time series analysis, the series of interest, denoted by the `key` variable, typically persist, through the trend modeling and also forecasting. We would typically want to examine the series across all of the keys.

Figure \ref{fig:highlight-retail} illustrates examining temporal data with many keys. The data has 152 series coorresponding to different items in retail data. The multiple series are displayed using an overlaid time series plot, along with a scatterplot of two variables (trend versus seasonal strength) from feature space, where each series is represented by a dot. The feature space is computed using the `features()` function from \CRANpkg{feasts}, which summarises the original data for each series using various statistical features. This function along with other **tidyverts** functions is `tsibble`-aware, and outputs a table in a reduced form where each row corresponds to a series, which can be graphically displayed as in Figure \ref{fig:highlight-retail-2}.

@@

Figure \ref{fig:highlight-retail} has also been highlighted to focus on the one series with the strongest seasonality. To create this highlighting, one needs to first filter the interesting series from the features table, and join back to the original `tsibble` in order to examine its trend in relation to others. This procedure can soon grow cumbersome if many series are to be explored. It illustrates a need to query interesting series on the fly. Although these two plots are static, we can consider them as linked views becaause the common `key` variables link between the two data tables producing the two ploots. This motivates the work in this package, described in this paper, to enable interactivity of `tsibble` and `tsibble`-derived objects for rapid exploratory data analysis.

### 2020-09-23 tsibbletalk.tex
background section

The \CRANpkg{tsibble} package \citep{wang2020tsibble} introduces a
unified temporal data structure, referred to as a \texttt{tsibble}, to
represent time series and longitudinal data in a tidy format
\citep{wickham2014tidy}. That said, a \texttt{tsibble} extends the
\texttt{data.frame} and \CRANpkg{tibble} classes with temporally
contextual metadata: \texttt{index} and \texttt{key}. The \texttt{index}
declares a data column that holds time-related indices. The \texttt{key}
identifies a collection of related series or panels observed over the
\texttt{index}-defined period, which can comprise multiple columns.
Below displays the monthly Australian retail trade turnover data
(\texttt{aus\_retail}), available in the \CRANpkg{tsibbledata} package
\citep{R-tsibbledata}. The \texttt{Month} column holds year-months as
\texttt{index}. The \texttt{State} together with \texttt{Industry} are
the identifiers for these 152 series, highlighted as \texttt{key}. Note
that the column \texttt{Series\ ID} could be an alternative option for
setting up \texttt{key}, but \texttt{State} and \texttt{Industry} are
more readable and informative. The \texttt{index} and \texttt{key} are
``sticky'' columns to a \texttt{tsibble}, forming critical pieces for
fluent temporal data analysis later.

@@

In the spirit of tidy data to the \CRANpkg{tidyverse}
\citep{Wickham2019}, the \textbf{tidyverts} suite features
\texttt{tsibble} as the foundational data structure, in order to build a
fluid and fluent pipeline for time series analysis. Besides
\CRANpkg{tsibble}, the \CRANpkg{feasts} \citep{R-feasts} and
\CRANpkg{fable} \citep{R-fable} packages fill the role of statistical
analysis and forecasting in the \textbf{tidyverts} ecosystem. When time
series analysis starts taking off, series of interest denoted by the
\texttt{key} variables often remain unchanged over the course of
analysis, from trend inspection to forecasting performance.

Figure \ref{fig:highlight-retail-1} gives an overview of 152 series for
the retail data using an overlaid time series plot, while Figure
\ref{fig:highlight-retail-2} presents a scatterplot, where each series
is represented by a dot in the feature space (trend versus seasonal
strength). The plot making of Figure \ref{fig:highlight-retail-2} is
aided with the \texttt{features()} function from \CRANpkg{feasts}, which
summarises original data by each series down to various statistical
features. This function along with other \textbf{tidyverts} functions is
\texttt{tsibble}-aware, and outputs a table in a reduced form where each
row corresponds to a series, thus graphically displayed as Figure
\ref{fig:highlight-retail-2}.

@@

Figure \ref{fig:highlight-retail} highlights not only a series with
strongest seasonality, but also a need to querying interesting series on
the fly. Without interactivity, one needs to first filter the
interesting series out from the features table, and join back to the
original \texttt{tsibble} in order to examine its trend in relation to
others. This procedure can soon grow cumbersome if many series to be
discovered. Despite that the two plots are static, they can be
considered as linked views via the common \texttt{key} variables between
two tables. This motivates enabling interactivity of \texttt{tsibble}
and \texttt{tsibble}-derived objects for rapid exploratory data
analysis.

>>>>

The \CRANpkg{tsibble} package \citep{wang2020tsibble} introduced a
unified temporal data structure, referred to as a \texttt{tsibble}, to
represent time series and longitudinal data in a tidy format
\citep{wickham2014tidy}. A \texttt{tsibble} extends the
\texttt{data.frame} and \CRANpkg{tibble} classes with the temporal
contextual metadata: \texttt{index} and \texttt{key}. The \texttt{index}
declares a data column that holds time-related indices. The \texttt{key}
identifies a collection of related series or panels observed over the
\texttt{index}-defined period, which can comprise multiple columns. An
example of a \texttt{tsibble} can be found in the monthly Australian
retail trade turnover data (\texttt{aus\_retail}), available in the
\CRANpkg{tsibbledata} package \citep{R-tsibbledata}, shown below. The
\texttt{Month} column holds year-months as the \texttt{index}.
\texttt{State} and \texttt{Industry}, together, are the identifiers for
these 152 series, which form the \texttt{key}. Note that the column
\texttt{Series\ ID} could be an alternative option for setting up the
\texttt{key}, but \texttt{State} and \texttt{Industry} are more readable
and informative. The \texttt{index} and \texttt{key} are ``sticky''
columns to a \texttt{tsibble}, forming critical pieces for fluent
downstream temporal data analysis.

@@

In the spirit of tidy data from the \CRANpkg{tidyverse}
\citep{Wickham2019}, the \textbf{tidyverts} suite features
\texttt{tsibble} as the foundational data structure, and helps to build
a fluid and fluent pipeline for time series analysis. Besides
\CRANpkg{tsibble}, the \CRANpkg{feasts} \citep{R-feasts} and
\CRANpkg{fable} \citep{R-fable} packages fill the role of statistical
analysis and forecasting in the \textbf{tidyverts} ecosystem. During all
the steps of a time series analysis, the series of interest, denoted by
the \texttt{key} variable, typically persist, through the trend modeling
and also forecasting. We would typically want to examine the series
across all of the keys.

Figure \ref{fig:highlight-retail} illustrates examining temporal data
with many keys. The data has 152 series coorresponding to different
items in retail data. The multiple series are displayed using an
overlaid time series plot, along with a scatterplot of two variables
(trend versus seasonal strength) from feature space, where each series
is represented by a dot. The feature space is computed using the
\texttt{features()} function from \CRANpkg{feasts}, which summarises the
original data for each series using various statistical features. This
function along with other \textbf{tidyverts} functions is
\texttt{tsibble}-aware, and outputs a table in a reduced form where each
row corresponds to a series, which can be graphically displayed as in
Figure \ref{fig:highlight-retail-2}.

@@

Figure \ref{fig:highlight-retail} has also been highlighted to focus on
the one series with the strongest seasonality. To create this
highlighting, one needs to first filter the interesting series from the
features table, and join back to the original \texttt{tsibble} in order
to examine its trend in relation to others. This procedure can soon grow
cumbersome if many series are to be explored. It illustrates a need to
query interesting series on the fly. Although these two plots are
static, we can consider them as linked views becaause the common
\texttt{key} variables link between the two data tables producing the
two ploots. This motivates the work in this package, described in this
paper, to enable interactivity of \texttt{tsibble} and
\texttt{tsibble}-derived objects for rapid exploratory data analysis.
